Using an iterated block particle filter via \pkg{spatPomp}

### 2022-12-12 bm.tex
coyping over from ibpf-paper/bm

\documentclass[12pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{url} 
\newcommand\eic[1]{{\textcolor{orange}{#1}}}
\newcommand\pnc[1]{{\textcolor{blue}{#1}}}
\usepackage{fullpage}
\bibliographystyle{apalike}

@@

<<cores,echo=F,cache=F,message = FALSE, warnings = FALSE,results='hide'>>=
library(doParallel)
cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()
registerDoParallel(cores)
ggplot2::theme_set(ggplot2::theme_bw())
@

<<packages,include=F,echo=F,cache=F>>=
library("spatPomp")
library("gridExtra")
library("grid")
library("ggplot2")
library("xtable")
library("tidyverse")
library("cowplot")
library("knitr")
library("kableExtra")
stopifnot(packageVersion("pomp")>="4.1")
@

<<set-opts,include=F,cache=F,purl=F>>=

options(
        scipen=2,
        help_type="html",
        stringsAsFactors=FALSE,
        prompt="R> ",
        continue="+  ",
        width=70,
        useFancyQuotes=FALSE,
        reindent.spaces=2,
        xtable.comment=FALSE
        )

@

<<knitr-opts,include=F,cache=F,purl=F>>=
opts_chunk$set(
    progress=TRUE,prompt=TRUE,highlight=FALSE,
#    tidy=TRUE,
#    tidy.opts=list(
#        keep.blank.line=FALSE
#    ),
    comment="",
    warning=FALSE,
    message=FALSE,
    error=TRUE,
    echo=TRUE,
    cache=FALSE,  ### SET CACHE TO TRUE WHILE EDITING TEXT
#    cache=TRUE,  ### SET CACHE TO FALSE WHILE DEBUGGING CODE
    strip.white=TRUE,
    results="markup",
    background="#FFFFFF00",
    size="normalsize",
    fig.path="figure/",
    fig.lp="fig:",
    fig.align="left",
    fig.show="asis",
    fig.height=5,fig.width=8,
    out.width="\\textwidth",
    dpi=300,
    dev="pdf",
    dev.args=list(
        bg="transparent",
        pointsize=12
    )
)
@

<<setup,cache=F,include=F>>=
myround<- function (x, digits = 1) {
  # adapted from the broman package
  # solves the bug that round() kills significant trailing zeros
  if (length(digits) > 1) {
    digits <- digits[1]
    warning("Using only digits[1]")
  }
  if (digits < 1) {
    as.character(round(x,digits))
  } else {
    tmp <- sprintf(paste("%.", digits, "f", sep = ""), x)
    zero <- paste0("0.", paste(rep("0", digits), collapse = ""))
    tmp[tmp == paste0("-", zero)] <- zero
    tmp
  }
}
mysignif <- function (x, digits = 1) {
  myround(x, digits - ceiling(log10(abs(x))))
}
@

\title{Using an iterated block particle filter via \pkg{spatPomp}}
\author{Ning Ning and Edward L. Ionides}
\date{\today}
\begin{document}
\maketitle

@@

<<bm-setup,echo=F,eval=T>>=
set.seed(20)
@

%%%%%% iiiiiiii %%%%
<<bm-model,echo=T,eval=T>>=
library(spatPomp)
i <- 2
b <- bm2(U=4,N=switch(i,20,200),unit_specific_names="rho")
plot(b)
@

<<bm-followup,echo=F,eval=T>>=
U <- length(unit_names(b))
@

Here, \code{i} is a computational intensity switch which adjusts the code for varying run-time.
We set \code{i=1} for testing and debugging, and \code{i=2} for higher quality results.
For simplicity, we consider only one unit-specific parameter, $\rho_{\unit}$, with other parameters being fixed at a value shared between units.
The simulation for \code{b} has $\rho_u=\Sexpr{coef(b)["rho1"]}$ for all $\unit$, but the estimators do not know this.
Before carrying out inference, we check likelihood evaluation.
For this toy model, the \pkg{spatPomp} function \code{bm2\_kalman\_logLik} provides an exact log-likelihood via the Kalman filter.
This study uses a sufficiently small number of units (\code{U=\Sexpr{U}}) that the particle filter is numerically tractable.
We use the particle filter provided by the \pkg{pomp} package \citep{king16}, taking advantage of the class structure where \class{spatPomp} inherits from \class{pomp}.
We can readily validate the agreement between \code{bm2\_kalman\_logLik} and \code{pfilter}, and identify the likelihood cost of the block filter approximation in this situation.

<<bm_logLik_code,cache=F,echo=T,eval=F>>=
kf_logLik <- bm2_kalman_logLik(b)
pf_logLik <- replicate(10,
  logLik(pfilter(b,switch(i,10,1000)))
)
bpf_logLik2 <- replicate(10,
  logLik(bpfilter(b,switch(i,10,1000),block_size=2))
)
@

<<bm_logLik_eval,cache=F,echo=F,eval=T>>=
out_dir <- paste0("out_",i,"/")
if(!dir.exists(out_dir)) dir.create(out_dir)
stew(file=paste0(out_dir,"logLik.rda"),seed=10,{
<<bm_logLik_code>>
  bpf_logLik1 <- replicate(10,
    logLik(bpfilter(b,switch(i,10,1000),block_size=1))
  )
  bpf_logLik4 <- replicate(10,
    logLik(bpfilter(b,switch(i,10,1000),block_size=4))
  )
})
@

@@

\\
\hline
Log-likelihood & mean &
  \Sexpr{myround(kf_logLik,2)} &
  \Sexpr{myround(mean(pf_logLik),2)} &
  \Sexpr{myround(mean(bpf_logLik4),2)} &
  \Sexpr{myround(mean(bpf_logLik2),2)} &
  \Sexpr{myround(mean(bpf_logLik1),2)} 
\\  
               & sd &
  0.00 &
  \Sexpr{myround(sd(pf_logLik),2)} &
  \Sexpr{myround(sd(bpf_logLik4),2)} &
  \Sexpr{myround(sd(bpf_logLik2),2)} &
  \Sexpr{myround(sd(bpf_logLik1),2)} 
\end{tabular}
\caption{Likelihood evaluation for the \code{bm2} model object, \code{b}, using the Kalman filter (KF), particle filter (PF), and block particle filter (BPF) with varying numbers of blocks ($K$).
For Monte Carlo filters, the mean and standard deviation are shown for 10 replicates.
}\label{tab:lik-eval}
\end{table}

@@

<<kf_mle,echo=F,eval=T>>=
stew(file=paste0(out_dir,"kf_mle.rda"),seed=999,{
  bm2_negLogLik <- function(rho){
    coef(b,names(rho)) <- unname(rho)
    -bm2_kalman_logLik(b)
  }

  rho_init <- c(rho1=0.4,rho2=0.4,rho3=0.4,rho4=0.4)
  bm2_negLogLik(rho_init)
  bm2_mle <- optim(rho_init,bm2_negLogLik)
})
@

<<kf_prof,echo=F,eval=T>>=
stew(file=paste0(out_dir,"kf_prof.rda"),seed=999,{
bm2_negProf <- function(rho_est,rho_fixed,u){
  rho_new <- rep(NA,U)
  names(rho_new) <- paste0("rho",1:U)
  rho_new[names(rho_est)] <- unname(rho_est)
  rho_new[u] <- rho_fixed
  coef(b,names(rho_new)) <- unname(rho_new)
  -bm2_kalman_logLik(b)
}

prof_init <- rep(0.4,U)
names(prof_init) <- paste0("rho",1:U)
prof_vals <- seq(from=0,to=0.8,length=switch(i,5,20))

kf_prof <- function(rho_vals,u){
  foreach(rho_prof=rho_vals)%dopar% {
    optim(prof_init[-u],bm2_negProf,rho_fixed=rho_prof,u=u)
  }
}

prof1 <- kf_prof(prof_vals,u=1)
prof2 <- kf_prof(prof_vals,u=2)
prof3 <- kf_prof(prof_vals,u=3)
prof4 <- kf_prof(prof_vals,u=4)

prof1_logLik <- sapply(prof1,function(x)-x$value)
prof2_logLik <- sapply(prof2,function(x)-x$value)
prof3_logLik <- sapply(prof3,function(x)-x$value)
prof4_logLik <- sapply(prof4,function(x)-x$value)
})

prof_range <- range(c(-bm2_mle$value,-bm2_mle$value-10))
if(0){
  par(mfrow=c(2,2))
  plot(y=prof1_logLik,x=prof_vals,ty="l",ylim=prof_range,
    ylab="Log-likelihood",xlab="rho1")
  plot(y=prof2_logLik,x=prof_vals,ty="l",ylim=prof_range,
    ylab="Log-likelihood",xlab="rho2")
  plot(y=prof3_logLik,x=prof_vals,ty="l",ylim=prof_range,
    ylab="Log-likelihood",xlab="rho3")
  plot(y=prof4_logLik,x=prof_vals,ty="l",ylim=prof_range,
    ylab="Log-likelihood",xlab="rho4")
}
@

<<ibpf-mle-code,eval=F,echo=T>>=
rho_start <- seq(from=0.2,to=0.8,length=U)
params_start <- coef(b)
params_start[paste0("rho",1:U)] <- rho_start
ibpf_mle_searches <- foreach(reps=1:switch(i,3,10))%dopar%{
  ibpf(b,params=params_start,
    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
    rw.sd=rw.sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
    unitParNames="rho",
    sharedParNames=NULL,
    block_size=2,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
@

<<ibpf-mle-eval,eval=T,echo=F>>=
stew(file=paste0(out_dir,"ibpf_mle.rda"),seed=999,{
<<ibpf-mle-code>>
})

<<ibpf-mle-lik,eval=T,echo=F>>=
stew(file=paste0(out_dir,"ibpf_mle_lik.rda"),seed=888,{
ibpf_kf_eval <- sapply(ibpf_mle_searches,bm2_kalman_logLik)
ibpf_bpf_eval <- foreach(bm2fit=ibpf_mle_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
  )
}
})
@

To assess the success of these searches, we evaluate the likelihood of the resulting parameter estimates using the Kalman filter.
The highest likelihood found in these ten searches was \Sexpr{myround(max(ibpf_kf_eval),2)}
 which is not far from the actual maximum of \Sexpr{myround(-bm2_mle$value,2)}.
However, the median of \Sexpr{myround(median(ibpf_kf_eval),2)} reveals that substantial Monte Carlo maximization error is present.
It can be intractable to increase computational effort to the point where the Monte Carlo error is negligible, and instead we emphasize methods that quantify and control this.

@@

<<ibpf-prof,eval=T,echo=F>>=
stew(file=paste0(out_dir,"prof1.rda"),seed=722,{
tic <- Sys.time()
u <- 1
rho_start <- seq(from=0.2,to=0.8,length=U)
reps <- switch(i,3,10)
param_start_matrix <- matrix(params_start,byrow=TRUE,
  ncol=length(coef(b)),nrow=reps*length(prof_vals),
  dimnames=list(NULL,names(coef(b))))
param_start_matrix[,paste0("rho",u)] <- rep(prof_vals,each=reps)
ibpf_prof_searches <- foreach(s=1:nrow(param_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("rho",x))
  ibpf(b,
    params=param_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="rho",
    sharedParNames=NULL,
    block_size=2,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
prof1time <- toc-tic
@

<<ibpf-prof-lik,eval=T,echo=F>>=
stew(file=paste0(out_dir,"prof1_lik.rda"),seed=868,{
tic <- Sys.time()
prof1_kf_eval <- sapply(ibpf_prof_searches,bm2_kalman_logLik)
ibpf_bpf_eval <- foreach(bm2fit=ibpf_prof_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
  )
}
ibpf_pf_eval <- foreach(bm2fit=ibpf_prof_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(pfilter(bm2fit,Np=switch(i,20,1000)))
  )
}
toc <- Sys.time()
})
prof1eval_time <- toc-tic
@

The profile took \Sexpr{myround(as.double(prof1time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
Likelihood evaluation took \Sexpr{myround(as.double(prof1eval_time,units="mins"),2)}, shared between BPF and PF.

<<ibpf-prof1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\rho_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=

par(mfrow=c(3,1))
par(mai=c(0.9,0.9,0.1,0.2))

##### bpf evaluation
p1bpf <- apply(matrix(apply(ibpf_bpf_eval,2,mean),nrow=reps),2,max)
bpf_prof_range <- c(max(p1bpf)-10,max(p1bpf))
plot(y=p1bpf,x=prof_vals,ylim=bpf_prof_range,
  ylab="Log-likelihood")
p1bpf_mcap <- mcap(logLik=p1bpf,parameter=prof_vals)
lines(p1bpf_mcap$fit$parameter,p1bpf_mcap$fit$smoothed)

abline(v=p1bpf_mcap$ci)

###### kf evaluation
p1kf <- apply(matrix(prof1_kf_eval,nrow=reps),2,max)
plot(y=p1kf,x=prof_vals,ylim=prof_range,
  ylab="Log-likelihood",xlab="")
p1kf_mcap <- mcap(logLik=p1kf,parameter=prof_vals)
lines(p1kf_mcap$fit$parameter,p1kf_mcap$fit$smoothed)
p1exact_mcap <- mcap(logLik=prof1_logLik,prof_vals)
# points(y=prof1_logLik,x=prof_vals,col="red")
lines(p1exact_mcap$fit$parameter,p1exact_mcap$fit$smoothed,col="red",lty="dashed")

abline(v=p1kf_mcap$ci)
abline(v=p1exact_mcap$ci,col="red",lty="dashed")

######## pf evaluation
p1pf <- apply(matrix(apply(ibpf_pf_eval,2,mean),nrow=reps),2,max)
pf_prof_range <- c(max(p1pf)-10,max(p1pf))
plot(y=p1pf,x=prof_vals,ylim=pf_prof_range,
  ylab="Log-likelihood",xlab="rho1")
p1pf_mcap <- mcap(logLik=p1pf,parameter=prof_vals)
lines(p1pf_mcap$fit$parameter,p1pf_mcap$fit$smoothed)

abline(v=p1pf_mcap$ci)

@

@@

<<bm-sigma-model,echo=T,eval=T>>=
set.seed(20)
b_sig <- bm2(U=4,N=switch(i,20,200),unit_specific_names="sigma")
# b_sig has the same data as b
# plot(b_sig)

  bm2_sig_negLogLik <- function(sigma){
    coef(b_sig,names(sigma)) <- unname(sigma)
    -bm2_kalman_logLik(b_sig)
  }

stew(file=paste0(out_dir,"kf_mle_sig.rda"),seed=256,{
  sigma_init <- c(sigma1=1,sigma2=1,sigma3=1,sigma4=1)
  bm2_sig_negLogLik(sigma_init)
  bm2_sig_mle <- optim(sigma_init,bm2_sig_negLogLik)
})

@

<<kf-prof-sig,echo=F,eval=T>>=
stew(file=paste0(out_dir,"kf_prof_sig.rda"),seed=512,{

bm2_sig_negProf <- function(sigma_est,sigma_fixed,u){
  sigma_new <- rep(NA,U)
  names(sigma_new) <- paste0("sigma",1:U)
  sigma_new[names(sigma_est)] <- unname(sigma_est)
  sigma_new[u] <- sigma_fixed
  coef(b_sig,names(sigma_new)) <- unname(sigma_new)
  -bm2_kalman_logLik(b_sig)
}

prof_sig_init <- rep(1,U)
names(prof_sig_init) <- paste0("sigma",1:U)
prof_sig_vals <- seq(from=0.4,to=1.6,length=switch(i,5,20))

kf_sig_prof <- function(sigma_vals,u){
  foreach(sigma_prof=sigma_vals)%dopar% {
    optim(prof_sig_init[-u],bm2_sig_negProf,sigma_fixed=sigma_prof,u=u)
  }
}

prof_sig1 <- kf_sig_prof(prof_sig_vals,u=1)
prof_sig2 <- kf_sig_prof(prof_sig_vals,u=2)
prof_sig3 <- kf_sig_prof(prof_sig_vals,u=3)
prof_sig4 <- kf_sig_prof(prof_sig_vals,u=4)

prof_sig1_logLik <- sapply(prof_sig1,function(x)-x$value)
prof_sig2_logLik <- sapply(prof_sig2,function(x)-x$value)
prof_sig3_logLik <- sapply(prof_sig3,function(x)-x$value)
prof_sig4_logLik <- sapply(prof_sig4,function(x)-x$value)
})

prof_sig_range <- range(c(-bm2_sig_mle$value,-bm2_sig_mle$value-10))
if(0){
  par(mfrow=c(2,2))
  plot(y=prof_sig1_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
    ylab="Log-likelihood",xlab="sigma1")
  plot(y=prof_sig2_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
    ylab="Log-likelihood",xlab="sigma2")
  plot(y=prof_sig3_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
    ylab="Log-likelihood",xlab="sigma3")
  plot(y=prof_sig4_logLik,x=prof_sig_vals,ty="l",ylim=prof_sig_range,
    ylab="Log-likelihood",xlab="sigma4")
}
@

<<ibpf-prof-sig,eval=T,echo=F>>=
stew(file=paste0(out_dir,"prof_sig1.rda"),seed=722,{
tic <- Sys.time()
u <- 1
sigma_start <- seq(from=0.4,to=1.6,length=U)
params_sig_start <- coef(b_sig)
params_sig_start[paste0("sigma",1:U)] <- sigma_start
reps <- switch(i,3,10)
param_sig_start_matrix <- matrix(params_sig_start,byrow=TRUE,
  ncol=length(coef(b_sig)),nrow=reps*length(prof_sig_vals),
  dimnames=list(NULL,names(coef(b_sig))))
param_sig_start_matrix[,paste0("sigma",u)] <- rep(prof_sig_vals,each=reps)
ibpf_prof_sig_searches <- foreach(s=1:nrow(param_sig_start_matrix))%dopar%{
  rw.sd_call <- rep(list(0.02),times=U-1)
  names(rw.sd_call) <- lapply((1:U)[-u],function(x)paste0("sigma",x))
  ibpf(b_sig,
    params=param_sig_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,20,1000),
    rw.sd=do.call(rw.sd,rw.sd_call),
    unitParNames="sigma",
    sharedParNames=NULL,
    block_size=2,
    spat_regression=0,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
prof_sig1time <- toc-tic
@

<<ibpf-prof-sig-lik,eval=T,echo=F>>=
stew(file=paste0(out_dir,"prof_sig1_lik.rda"),seed=868,{
tic <- Sys.time()
prof_sig1_kf_eval <- sapply(ibpf_prof_sig_searches,bm2_kalman_logLik)
ibpf_bpf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(bpfilter(bm2fit,block_size=2,Np=switch(i,20,1000)))
  )
}
ibpf_pf_sig_eval <- foreach(bm2fit=ibpf_prof_sig_searches,.combine=cbind) %dopar% {
  replicate(switch(i,5,20),
    logLik(pfilter(bm2fit,Np=switch(i,20,1000)))
  )
}
toc <- Sys.time()
})
prof_sig1eval_time <- toc-tic
@

The $\sigma_1$ profile took \Sexpr{myround(as.double(prof_sig1time,units="mins"),2)} mins using \Sexpr{cores} computing cores.
Likelihood evaluation took \Sexpr{myround(as.double(prof_sig1eval_time,units="mins"),2)}, shared between BPF and PF.

<<ibpf-prof-sig1-plot, fig.height=8, fig.width=6, out.width="5.5in", fig.cap = "Top: profile for $\\sigma_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter.",eval=T,echo=F>>=

par(mfrow=c(3,1))
par(mai=c(0.9,0.9,0.1,0.2))

##### bpf evaluation
p1_sig_bpf <- apply(matrix(apply(ibpf_bpf_sig_eval,2,mean),nrow=reps),2,max)
bpf_prof_sig_range <- c(max(p1_sig_bpf)-10,max(p1_sig_bpf))
plot(y=p1_sig_bpf,x=prof_sig_vals,ylim=bpf_prof_sig_range,
  ylab="Log-likelihood")
p1_sig_bpf_mcap <- mcap(logLik=p1_sig_bpf,parameter=prof_sig_vals)
lines(p1_sig_bpf_mcap$fit$parameter,p1_sig_bpf_mcap$fit$smoothed)

abline(v=p1_sig_bpf_mcap$ci)

###### kf evaluation
p1_sig_kf <- apply(matrix(prof_sig1_kf_eval,nrow=reps),2,max)
plot(y=p1_sig_kf,x=prof_sig_vals,ylim=prof_sig_range,
  ylab="Log-likelihood",xlab="")
p1_sig_kf_mcap <- mcap(logLik=p1_sig_kf,parameter=prof_sig_vals)
lines(p1_sig_kf_mcap$fit$parameter,p1_sig_kf_mcap$fit$smoothed)
p1_sig_exact_mcap <- mcap(logLik=prof_sig1_logLik,prof_sig_vals)
# points(y=prof_sig1_logLik,x=prof_sig_vals,col="red")
lines(p1_sig_exact_mcap$fit$parameter,p1_sig_exact_mcap$fit$smoothed,col="red",lty="dashed")

abline(v=p1_sig_kf_mcap$ci)
abline(v=p1_sig_exact_mcap$ci,col="red",lty="dashed")

######## pf evaluation
p1_sig_pf <- apply(matrix(apply(ibpf_pf_sig_eval,2,mean),nrow=reps),2,max)
pf_prof_sig_range <- c(max(p1_sig_pf)-10,max(p1_sig_pf))
plot(y=p1_sig_pf,x=prof_sig_vals,ylim=pf_prof_sig_range,
  ylab="Log-likelihood",xlab="sigma1")
p1_sig_pf_mcap <- mcap(logLik=p1_sig_pf,parameter=prof_sig_vals)
lines(p1_sig_pf_mcap$fit$parameter,p1_sig_pf_mcap$fit$smoothed)

abline(v=p1_sig_pf_mcap$ci)

@

@@

<<model-he10-code,eval=F,echo=T>>=
he10_model <- he10(U=4,dt=1/365,Tmax=1964,expandedParNames=c("R0"),
  basic_params = c(
    alpha =0.99,      iota=0,          R0=30,
    cohort=0.5,  amplitude=0.3,     gamma=52,
    sigma=52,           mu=0.02,  sigmaSE=0.05,
    rho=0.5,           psi=0.1,         g=1000,
    S_0=0.036,         E_0=0.00007,   I_0=0.00006
  )
)
m <- simulate(he10_model,seed=9)
@

<<model-he10-eval,eval=T,echo=F>>=
stew(file=paste0(out_dir,"m-sim.rda"),{
<<model-he10-code>>
})
plot(he10_model,log=T)
plot(m,log=T)
@

>>>>

\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{xcolor}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{soul}
\usepackage{xcolor}
\usepackage{url} 
\newcommand\eic[1]{{\textcolor{orange}{#1}}}
\newcommand\pnc[1]{{\textcolor{blue}{#1}}}
\usepackage{fullpage}
\bibliographystyle{apalike}

@@

\title{Using an iterated block particle filter via \pkg{spatPomp}}
\author{Ning Ning and Edward L. Ionides}
\date{\today}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle

@@

%%%%%% iiiiiiii %%%%
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
R> library(spatPomp)
R> i <- 2
R> b <- bm2(U=4,N=switch(i,20,200),unit_specific_names="rho")
R> plot(b)
\end{verbatim}
\end{kframe}

\includegraphics[width=\textwidth]{figure/bm-model-1} \hfill{}

\end{knitrout}

Here, \code{i} is a computational intensity switch which adjusts the code for varying run-time.
We set \code{i=1} for testing and debugging, and \code{i=2} for higher quality results.
For simplicity, we consider only one unit-specific parameter, $\rho_{\unit}$, with other parameters being fixed at a value shared between units.
The simulation for \code{b} has $\rho_u=0.4$ for all $\unit$, but the estimators do not know this.
Before carrying out inference, we check likelihood evaluation.
For this toy model, the \pkg{spatPomp} function \code{bm2\_kalman\_logLik} provides an exact log-likelihood via the Kalman filter.
This study uses a sufficiently small number of units (\code{U=4}) that the particle filter is numerically tractable.
We use the particle filter provided by the \pkg{pomp} package \citep{king16}, taking advantage of the class structure where \class{spatPomp} inherits from \class{pomp}.
We can readily validate the agreement between \code{bm2\_kalman\_logLik} and \code{pfilter}, and identify the likelihood cost of the block filter approximation in this situation.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
R> kf_logLik <- bm2_kalman_logLik(b)
R> pf_logLik <- replicate(10,
+    logLik(pfilter(b,switch(i,10,1000)))
+  )
R> bpf_logLik2 <- replicate(10,
+    logLik(bpfilter(b,switch(i,10,1000),block_size=2))
+  )
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\\
\hline
Log-likelihood & mean &
  -1472.69 &
  -1480.79 &
  -1477.88 &
  -1509.06 &
  -1547.80 
\\  
               & sd &
  0.00 &
  3.96 &
  4.00 &
  2.07 &
  1.49 
\end{tabular}
\caption{Likelihood evaluation for the \code{bm2} model object, \code{b}, using the Kalman filter (KF), particle filter (PF), and block particle filter (BPF) with varying numbers of blocks ($K$).
For Monte Carlo filters, the mean and standard deviation are shown for 10 replicates.
}\label{tab:lik-eval}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
R> rho_start <- seq(from=0.2,to=0.8,length=U)
R> params_start <- coef(b)
R> params_start[paste0("rho",1:U)] <- rho_start
R> ibpf_mle_searches <- foreach(reps=1:switch(i,3,10))%dopar%{
+    ibpf(b,params=params_start,
+      Nbpf=switch(i,2,50),Np=switch(i,20,1000),
+      rw.sd=rw.sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
+      unitParNames="rho",
+      sharedParNames=NULL,
+      block_size=2,
+      spat_regression=0,
+      cooling.fraction.50=0.5
+    )
+  }
\end{verbatim}
\end{kframe}
\end{knitrout}

To assess the success of these searches, we evaluate the likelihood of the resulting parameter estimates using the Kalman filter.
The highest likelihood found in these ten searches was -1472.93
 which is not far from the actual maximum of -1472.11.
However, the median of -1475.00 reveals that substantial Monte Carlo maximization error is present.
It can be intractable to increase computational effort to the point where the Monte Carlo error is negligible, and instead we emphasize methods that quantify and control this.

@@

The profile took 56.11 mins using 10 computing cores.
Likelihood evaluation took 25.07, shared between BPF and PF.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}

\includegraphics[width=5.5in]{figure/ibpf-prof1-plot-1} \hfill{}

\caption[Top]{Top: profile for $\rho_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter.}\label{fig:ibpf-prof1-plot}
\end{figure}

\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
R> set.seed(20)
R> b_sig <- bm2(U=4,N=switch(i,20,200),unit_specific_names="sigma")
R> # b_sig has the same data as b
R> # plot(b_sig)
R> 
R>   bm2_sig_negLogLik <- function(sigma){
+      coef(b_sig,names(sigma)) <- unname(sigma)
+      -bm2_kalman_logLik(b_sig)
+    }
R> 
R> stew(file=paste0(out_dir,"kf_mle_sig.rda"),seed=256,{
+    sigma_init <- c(sigma1=1,sigma2=1,sigma3=1,sigma4=1)
+    bm2_sig_negLogLik(sigma_init)
+    bm2_sig_mle <- optim(sigma_init,bm2_sig_negLogLik)
+  })
\end{verbatim}
\end{kframe}
\end{knitrout}

The $\sigma_1$ profile took 55.87 mins using 10 computing cores.
Likelihood evaluation took 23.72, shared between BPF and PF.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{figure}

\includegraphics[width=5.5in]{figure/ibpf-prof-sig1-plot-1} \hfill{}

\caption[Top]{Top: profile for $\sigma_1$ using an IBPF search with likelihood computed using BPF, for $K=2$ blocks each having 2 units. Middle: Exact profile (dashed red line) and the same IBPF search with likelihood computed exactly using the Kalman filter. Bottom: The same IBPF search with likelihood computed using the particle filter.}\label{fig:ibpf-prof-sig1-plot}
\end{figure}

\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
R> he10_model <- he10(U=4,dt=1/365,Tmax=1964,expandedParNames=c("R0"),
+    basic_params = c(
+      alpha =0.99,      iota=0,          R0=30,
+      cohort=0.5,  amplitude=0.3,     gamma=52,
+      sigma=52,           mu=0.02,  sigmaSE=0.05,
+      rho=0.5,           psi=0.1,         g=1000,
+      S_0=0.036,         E_0=0.00007,   I_0=0.00006
+    )
+  )
R> m <- simulate(he10_model,seed=9)
\end{verbatim}
\end{kframe}
\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}

\includegraphics[width=\textwidth]{figure/model-he10-eval-1} \hfill{}

\includegraphics[width=\textwidth]{figure/model-he10-eval-2} \hfill{}

\end{knitrout}

Likelihood evaluation took 23.72 mins.
